---
title: "selection_coefficient"
author: "ALARCON Robin"
date: "2026-05-07"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
editor_options: 
  chunk_output_type: console
---
# set-up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,  
                      message = FALSE,
                      warning = FALSE)

setwd('/Users/robinalarcon/Desktop/stage IBENS/direct selection')
```

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(lme4)
library(lmerTest)
library(ggh4x)
library(readxl)
library(ggsignif)
library(stringr)
```

# charging data

```{r}
dat <- read.csv("predicted_frequencies_block2.csv", sep = ";") # 653 obs
pop_info <- read_xlsx("pop_information.xlsx")
```

## data reformatting


```{r}
sample <- dat %>%
  filter(is.na(calib))%>% # 473obs
  filter(!is.na(predicted_freq)) %>%# 413 obs
  filter(normEBS.decrease != 0) %>% #wells for which amplification and melting failed are filtered out
  dplyr::select(sample,pop, setup, generation, competition, predicted_freq, experiment_batch, plate_qpcr_id)%>%
  mutate(comp_back = case_when(competition == "V5"~"ECA722",
                               competition == "V6"~"ECA191",
                               competition == "V7"~"ECA730",
                               competition == "V8"~"ECA723")) ## data reformatting, and genetic backgrounds added as a variable (column "comp_back") based on the working code in the column "competition"
```


```{r}
# all data for plots
setup_map <- pop_info[1:48,c(1,4)]

setup <- sample %>%
  filter(generation == 0) %>%
  left_join(setup_map, by = "setup")%>%
  dplyr::select(-pop.x)%>%
  dplyr::rename("pop"="pop.y")%>%
  dplyr::relocate(pop, .before = 1)

non_setup <- sample %>%
  filter(generation != 0)

all_data <- bind_rows(non_setup, setup) %>%
  arrange(pop, generation)%>%
  mutate( envi = case_when( str_detect(pop, "Z") ~ "ZINC",
                           TRUE ~ "NGM")) # environment is added as a new variable, by detecting the prefix "Z" in the population identifier

# mean data
mean_data <- all_data %>%
  group_by(pop, generation, setup, competition, comp_back, experiment_batch, envi) %>%
  dplyr::summarise(
    mean_freq = mean(predicted_freq, na.rm = TRUE) #mean of the 3 technical replicates for all population/generation pairs (other variables are kept in the new mean_data, such as the setup lineage)
  )

```




# selection coefficient


## logit transformation

```{r}
library(RColorBrewer)

logit_data <- all_data %>%
  mutate(logit_freq = qlogis(predicted_freq))
logit_mean <- mean_data %>%
  mutate(logit_mean_freq = qlogis(mean_freq))


logit_dataNGM <- logit_data %>% filter(envi == "NGM")
logit_dataZINC <- logit_data %>% filter(envi == "ZINC")

logit_meanNGM <- logit_mean %>% filter(envi == "NGM")
logit_meanZINC <- logit_mean %>% filter(envi == "ZINC")

colorsBRBG <- brewer.pal(n = 4, name = "BrBG")
print(colorsBRBG)

colorsBRBGselected <- colorsBRBG[c(1,4)]

install.packages("ggpubr")
install.packages("patchwork")
library(ggpubr)
library(patchwork)

#separate plots built for NGM and Zinc
pNGM <- ggplot() +
  scale_colour_manual(values = colorsBRBG, name = "genetic background") +
  geom_point(data = logit_dataNGM, 
             aes(x = generation, y = logit_freq, group = pop, color = comp_back),
             alpha = 0.2)+
  geom_line(data = logit_meanNGM, 
            aes(x = generation, y = logit_mean_freq, group = pop, color = comp_back),
            alpha = 0.4, linewidth = 1.5) +
  annotate("text", x = 5, y = 1, label = "NGM",
           color = "black", size = 10, alpha = 0.5) +
  annotate("text", x = 3, y = Inf, label = "(a)",
           vjust = 1.5, hjust = -0.1, size = 5, color = "black")+
  coord_cartesian(clip = "off") +
  theme_classic()+
  theme(
  axis.title = element_text(size = 15),
  axis.text = element_text(size = 12))+
  xlab("Generation") +
  ylab("logit rec-1 mutant frequency")

pZINC <- ggplot() +
  scale_colour_manual(values = colorsBRBG, name = "genetic background") +
  geom_point(data = logit_dataZINC, 
             aes(x = generation, y = logit_freq, group = pop, color = comp_back),
             alpha = 0.2)+
  geom_line(data = logit_meanZINC, 
            aes(x = generation, y = logit_mean_freq, group = pop, color = comp_back),
            alpha = 0.4, linewidth = 1.5) +
  annotate("text", x = 5, y = 1, label = "ZINC",
           color = "black", size = 10, alpha = 0.5) +
  annotate("text", x = 3, y = Inf, label = "(b)",
           vjust = 1.5, hjust = -0.1, size = 5, color = "black")+
  coord_cartesian(clip = "off") +
  theme_classic()+
  theme(
  axis.title = element_text(size = 15),
  axis.text = element_text(size = 12))+
  xlab("Generation") +
  ylab("logit rec-1 mutant frequency")


combined_plot <- ggarrange(pNGM, pZINC, ncol = 2, common.legend = TRUE)
print(combined_plot)


# 4 plots wrapped depending on the genetic background with a common caption showing the environment of evolution
ggplot(logit_mean, aes(x = generation, y = logit_mean_freq, group = pop, color = envi)) +
  scale_colour_manual(values = colorsBRBGselected, name = "environment") +
  facet_wrap(~comp_back)+
  geom_line(alpha = 0.5,linewidth = 1.2) +
  geom_point(data = logit_data, aes(x = generation, y = logit_freq, group = pop, color = envi), alpha = 0.2)+
  theme_classic()+
  theme(
  axis.title = element_text(size = 15),
  axis.text = element_text(size = 12))+
  xlab("Generation") +
  ylab("logit-transformed estimated rec-1 mutant frequency")
```


## check data distribution

```{r}
hist(x = mean_data$mean_freq, breaks = 30)
hist(x = logit_mean$logit_mean_freq, breaks = 30)
```

## first model

### model with nested effects similar to those of model from experiment 1

fixed effect : intercept, generation, envi, genetic background

random effect (on intercept and slope): pop

-> boundary (singular) fit: see help('isSingular')

-> meaning that model is too complexe (random effect structure)

-> corr for pop btw intercept and generation is -1

```{r}

logit_mean_no0 <- logit_mean %>% filter(generation != 0)
logit_data_0 <- logit_data %>% filter(generation == 0)

# variability in frequencies in founding populations is examined
ggplot()+
  geom_point(data = logit_data_0, 
             aes(x = comp_back, y = predicted_freq, group = pop, color=sample),
             alpha = 0.2) # a lot of variation around the 0.25 ratio that we aimed at !

mod_complex <- lmer(
  logit_mean_freq ~ 1 + generation*envi*comp_back + (generation | setup/pop ) + (1 | setup ),
  data = logit_mean
)
summary(mod_complex)

#too many interactions to interpret if we do 1 single model
#we'd rather do 1 model/genetic background

modECA191 <- lmer(
  logit_mean_freq ~ 1 + generation*envi - envi + (generation | setup/pop ) + (1 | setup ), #environment is removed from the fixed effects on the intercept, because otherwise it does not make sense in our design: independently of the environment, the intercept is determined by the setup of the specific population
  data = logit_mean %>% filter(comp_back == "ECA191")
)
summary(modECA191)

modECA722 <- lmer(
  logit_mean_freq ~ 1 + generation*envi - envi + (generation | setup/pop ) + (1 | setup ),
  data = logit_mean %>% filter(comp_back == "ECA722")
)
summary(modECA722)

modECA723 <- lmer(
  logit_mean_freq ~ 1 + generation*envi - envi + (generation | setup/pop ) + (1 | setup ),
  data = logit_mean %>% filter(comp_back == "ECA723")
)
summary(modECA723)

modECA730 <- lmer(
  logit_mean_freq ~ 1 + generation*envi -  envi + (generation | setup/pop ) + (1 | setup ),
  data = logit_mean %>% filter(comp_back == "ECA730")
)
summary(modECA730)

#for each of the 4 models, random effect of generation is strongly correlated to the random effect of the intercept. It is not improving the model in any way, so only the effect on the intercept is kept in the next chunk, because of the high variance observed in founding populations.

```

### second model, simplified

fixed effect : intercept, generation, envi, genetic background

random effect (on intercept): pop


```{r}

modECA191simple <- lmer(
  logit_mean_freq ~ 1 + generation*envi - envi + (1 | setup ),
  data = logit_mean %>% filter(comp_back == "ECA191")
)
summary(modECA191simple)

modECA722simple <- lmer(
  logit_mean_freq ~ 1 + generation*envi - envi + (1 | setup ),
  data = logit_mean %>% filter(comp_back == "ECA722")
)
summary(modECA722simple)

modECA723simple <- lmer(
  logit_mean_freq ~ 1 + generation*envi - envi + (1 | setup ),
  data = logit_mean %>% filter(comp_back == "ECA723")
)
summary(modECA723simple)

modECA730simple <- lmer(
  logit_mean_freq ~ 1 + generation*envi - envi + (1 | setup ),
  data = logit_mean %>% filter(comp_back == "ECA730")
)
summary(modECA730simple)
```
### potential experimenter effect

Anova P value >> 0.05 --> no significant influence of batch effect

```{r}
fit_batch730 <- lmer(
  logit_mean_freq ~ 1 + generation*envi - envi + (1 | setup ) + (generation | experiment_batch),
  data = logit_mean %>% filter(comp_back == "ECA730")
)

anova(modECA730simple, fit_batch730)

fit_batch723 <- lmer(
  logit_mean_freq ~ 1 + generation*envi - envi + (1 | setup ) + (generation | experiment_batch),
  data = logit_mean %>% filter(comp_back == "ECA723")
)

anova(modECA723simple, fit_batch723)

fit_batch722 <- lmer(
  logit_mean_freq ~ 1 + generation*envi - envi + (1 | setup ) + (generation | experiment_batch),
  data = logit_mean %>% filter(comp_back == "ECA722")
)

anova(modECA722simple, fit_batch722)

fit_batch191 <- lmer(
  logit_mean_freq ~ 1 + generation*envi - envi + (1 | setup ) + (generation | experiment_batch),
  data = logit_mean %>% filter(comp_back == "ECA191")
)

anova(modECA191simple, fit_batch191)
 # ANOVA is performed between our models and a model with an additional random effect of the experimenter on the slope.

```


### extracting slopes of the models, namely selection coefficients

```{r}
library(emmeans)

# extract generation effect (slope) for every pair of environment/background
slopesECA191 <- emtrends(
  modECA191simple,
  ~ envi,
  var = "generation"
)


slopesECA722 <- emtrends(
  modECA722simple,
  ~ envi,
  var = "generation"
)

slopesECA723 <- emtrends(
  modECA723simple,
  ~ envi,
  var = "generation"
)

slopesECA730 <- emtrends(
  modECA730simple,
  ~ envi,
  var = "generation"
)



selection_table_ECA191 <- summary(slopesECA191, infer = c(TRUE, TRUE)) %>%
  as.data.frame() %>% 
  dplyr::rename(
    selection_effect = generation.trend,
    SE = SE,
    lower_CI = lower.CL,
    upper_CI = upper.CL
  )
selection_table_ECA191 <- selection_table_ECA191 %>% mutate(background="ECA191")


selection_table_ECA722 <- summary(slopesECA722, infer = c(TRUE, TRUE)) %>%
  as.data.frame() %>% 
  dplyr::rename(
    selection_effect = generation.trend,
    SE = SE,
    lower_CI = lower.CL,
    upper_CI = upper.CL
  )
selection_table_ECA722 <- selection_table_ECA722 %>% mutate(background="ECA722")

selection_table_ECA723 <- summary(slopesECA723, infer = c(TRUE, TRUE)) %>%
  as.data.frame() %>% 
  dplyr::rename(
    selection_effect = generation.trend,
    SE = SE,
    lower_CI = lower.CL,
    upper_CI = upper.CL
  )
selection_table_ECA723 <- selection_table_ECA723 %>% mutate(background="ECA723")

selection_table_ECA730 <- summary(slopesECA730, infer = c(TRUE, TRUE)) %>%
  as.data.frame() %>% 
  dplyr::rename(
    selection_effect = generation.trend,
    SE = SE,
    lower_CI = lower.CL,
    upper_CI = upper.CL
  )
selection_table_ECA730 <- selection_table_ECA730 %>% mutate(background="ECA730")


selection_global <- bind_rows(selection_table_ECA191, selection_table_ECA722, selection_table_ECA723, selection_table_ECA730)


```



# Visualization of selection coefficients

## selection coef

```{r, fig.width=10, fig.height=5}

library(RColorBrewer)

colorsBRBG <- brewer.pal(n = 4, name = "BrBG")
print(colorsBRBG)

colorsBRBGselected <- colorsBRBG[c(2,4)]

ggplot(selection_global, aes(x = 1, y = selection_effect, color = envi)) +
  facet_wrap(~background) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_color_manual(values = colorsBRBGselected, name= "environment") +
  geom_point(position = position_dodge(width = 0.4), size = 1.7) +
  geom_errorbar(
    aes(ymin = lower_CI, ymax = upper_CI),
    position = position_dodge(width = 0.4),
    width = 0.03
  ) +
  scale_y_continuous(limits = c(-0.15, 0.15)) +
  scale_x_continuous(breaks = NULL) +  # Remove x-axis ticks and labels
  theme_classic() +
  xlab("") +  # Remove x-axis label
  ylab("Selection coefficient") +
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    axis.text.x = element_blank(),  # Remove x-axis text
    axis.ticks.x = element_blank(),  # Remove x-axis ticks
    axis.title.x = element_blank()   # Remove x-axis title
  )


```
