---
title: "selection_coefficient"
author: "ALARCON Robin"
date: "2026-05-07"
output:
  pdf_document: default
  html_document: default
editor_options:
  chunk_output_type: inline
---
# set-up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,  
                      message = FALSE,
                      warning = FALSE)

setwd('/Users/robinalarcon/Desktop/stage IBENS/direct selection')
```

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(lme4)
library(lmerTest)
library(ggh4x)
```

# data

```{r}
dat <- read.csv("predicted_frequencies_block1.csv", sep = ";")

sample <- dat %>%
  filter(is.na(calib))

# mean of predicted frequencies is calculated in mean_sample for each population at every generation
mean_sample <- sample %>%
  dplyr::group_by(pop, generation, mix, pop_id) %>%
  dplyr::summarise(
    mean_freq = mean(predicted_freq, na.rm = TRUE)
  )
```

# rec frequency plot

```{r}
# here we organize mean frequencies by population, pop_id, generation and mix (setup lineage used to seed founding population)
setup_map <- data.frame(
  pop_id = rep(LETTERS[1:8], each = 4),
  pop_id_new = unlist(lapply(LETTERS[1:8], function(x) {
    paste0(x, sprintf("%02d", 1:4))
  }))
)
setup_expanded <- mean_sample %>%
  filter(generation == 0) %>%
  left_join(setup_map, by = "pop_id") %>%
  mutate(pop_id = pop_id_new) %>%
  dplyr::select(-pop_id_new)
non_setup <- mean_sample %>%
  filter(generation != 0)
mean_data <- bind_rows(non_setup, setup_expanded) %>%
  arrange(pop_id, generation)

# all data required for plotting, whith plate identifiers
setup_expanded_all <- sample %>%
  filter(generation == 0) %>%
  left_join(setup_map, by = "pop_id") %>%
  mutate(pop_id = pop_id_new) %>%
  dplyr::select(-pop_id_new)
non_setup_all <- sample %>%
  filter(generation != 0)
all_data <- bind_rows(non_setup_all, setup_expanded_all) %>%
  arrange(pop_id, generation)


```

```{r, fig.width=6, fig.height=4}
library(khroma)
library(RColorBrewer)

# variation of estimated rec-1 frequency is plotted against generation
ggplot(mean_data, aes(x = generation, y = mean_freq, group = pop_id, color = mix)) +
  geom_line(alpha = 0.5,linewidth = 1.2) +
  scale_color_brewer(palette="BrBG")+
  scale_x_continuous(
    breaks = c(0, 1, 4, 8)) +
  scale_y_continuous(
    limits = c(0.4, 0.8)) +
  theme_classic()+
  theme(
  axis.title = element_text(size = 15),
  axis.text = element_text(size = 12))+
  xlab("Generation") +
  ylab("Estimated rec-1 mutant frequency")
```


```{r, fig.height=3.5, fig.width=6}
library(RColorBrewer)
library(palettetown)

all_dataAD <- all_data %>%
  filter(mix %in% c("A", "B", "C", "D"))
all_dataEH <- all_data %>%
  filter(mix %in% c("E", "F", "G", "H"))

mean_dataAD <- mean_data %>%
  filter(mix %in% c("A", "B", "C", "D"))
mean_dataEH <- mean_data %>%
  filter(mix %in% c("E", "F", "G", "H"))

all_logitdataAD <- all_data %>%
  filter(mix %in% c("A", "B", "C", "D")) %>% 
  mutate(predicted_freq=qlogis(predicted_freq))
all_logitdataEH <- all_data %>%
  filter(mix %in% c("E", "F", "G", "H")) %>% 
  mutate(predicted_freq=qlogis(predicted_freq))

mean_logitdataAD <- mean_data %>%
  filter(mix %in% c("A", "B", "C", "D")) %>% 
  mutate(mean_freq=qlogis(mean_freq))
mean_logitdataEH <- mean_data %>%
  filter(mix %in% c("E", "F", "G", "H")) %>% 
  mutate(mean_freq=qlogis(mean_freq))


```


# frequencies are logit-transformed


## transformation to logit is the last step before estimating selection coefficient

```{r}
# all transformations to logit are perofrmed below
logit_data <- all_data %>% # we also transform all technical replicates to make all data points (and not just lines) appear on the following plots
  mutate(logit_freq = qlogis(predicted_freq))
logit_mean <- mean_data %>%
  mutate(logit_mean_freq = qlogis(mean_freq))

# here we build subsets of logit_data and logit_mean depending on the mix used to seed the founding population
logit_ADdata <- logit_data %>% 
  filter(mix %in% c("A", "B", "C", "D"))

logit_EHdata <- logit_data %>% 
  filter(mix %in% c("E", "F", "G", "H"))

logit_ADmean <- logit_mean %>% 
  filter(mix %in% c("A", "B", "C", "D"))

logit_EHmean <- logit_mean %>% 
  filter(mix %in% c("E", "F", "G", "H"))

# two plots corresponding to the mix subsets are built (they are present in the report as figure 5)
ggplot() +
  geom_point(data = logit_ADdata, 
             aes(x = generation, y = logit_freq, group = pop_id, color = mix),
             alpha = 0.2)+
  geom_line(data = logit_ADmean, 
            aes(x = generation, y = logit_mean_freq, group = pop_id, color = mix),
            alpha = 0.4, linewidth = 1.5) +
  scale_colour_brewer(palette = "BrBG", direction = -1)+
  theme_classic()+
  theme(
  axis.title = element_text(size = 15),
  axis.text = element_text(size = 12))+
  xlab("Generation") +
  ylab("logit-transformed rec-1 mutant frequency")

ggplot() +
  geom_point(data = logit_EHdata, 
             aes(x = generation, y = logit_freq, group = pop_id, color = mix),
             alpha = 0.2)+
  geom_line(data = logit_EHmean, 
            aes(x = generation, y = logit_mean_freq, group = pop_id, color = mix),
            alpha = 0.4, linewidth = 1.5) +
  scale_colour_brewer(palette = "BrBG", direction = -1)+
  theme_classic()+
  theme(
  axis.title = element_text(size = 15),
  axis.text = element_text(size = 12))+
  xlab("Generation") +
  ylab("logit-transformed rec-1 mutant frequency")
```


## check data distribution

```{r}
hist(x = mean_data$mean_freq, breaks = 30)
hist(x = logit_mean$logit_mean_freq, breaks = 30)
```

## fa first model is built with nested random effect on the slope and the intercept


fixed effect : intercept and generation

random effect (on intercept and slope): mix and population within mix

-> warning: Model failed to converge with max|grad| = 0.00403952 (tol = 0.002, component 1)

-> meaning that model is too complexe (random effect structure)

-> corr for pop_id:mix btw intercept and generation is -1

```{r}
install.packages("lme4")

library(Matrix)
library(lme4)

selection_complexe <- lmer(logit_mean_freq
   ~ 1+ generation + (generation | mix / pop_id ),
  data = logit_mean
)

summary(selection_complexe)
```

### a simplified model is built, respecting structure in the data
## nested random effect is kept for slope only

fixed effect : intercept + generation

random effect on intercept : mix

random  effect on slope : mix and pop within mix

```{r}

logit_mean2 <- logit_mean %>%
  mutate(mix=as.factor(mix)) %>% 
  mutate(pop_id=as.factor(pop_id))

selection_mod <- lmer(
  logit_mean_freq ~ 1+ generation + (0+generation| mix/pop_id ) + (1|mix), # random effect around intercept of each pop inside mix is removed. mix alone is expected to induce variance in the intercept, hence the random effect (1|mix).
  data = logit_mean2
)
summary(selection_mod)


#significance of random effects: bootstrapping

logit_mean2 <- na.omit(logit_mean2)

nBoot <- 1000
lrStat <- rep(NA, nBoot)
ft.null <- lm(logit_mean_freq ~ 1+ generation, data = logit_mean2) # null model
ft.alt <- selection_mod # alternate model
lrObs <- 2 * logLik(ft.alt) - 2 * logLik(ft.null) # observed test stat
for (iBoot in 1:nBoot)
{
  logit_mean2$freqSim <- unlist(simulate(ft.null)) # resampled data
  bNull <-lm(freqSim ~ 1+ generation, data = logit_mean2) # null model
  bAlt <- lmer(freqSim ~ 1+ generation + (0+generation| mix/pop_id ) + (1|mix),
  data = logit_mean2) # alternate model
  lrStat[iBoot] <- 2 * logLik(bAlt) - 2 * logLik(bNull) # resampled test stat
}
mean(lrStat > lrObs)#P-value for test of the random effect

lrObs

hist(lrStat, col='blue', xlim=c(-15, 148)) # Histogram of the 1000 values of likelihood of the simulated model
abline(v = lrObs, col="red", lwd=3, lty=2) # Vertical red line representing the likelihood of the observed model including the random factor

# test of the model's assumptions: normality of residuals and homogeneity of variance

#QQplot
library(DHARMa)

residuals_mod <- simulateResiduals(selection_mod)
plot(residuals_mod, main="QQ-plot of simulated residuals")

testDispersion(residuals_mod)



# residuals vs fitted
plot(residuals(selection_mod)~fitted(selection_mod)
      , col='blue'
      , pch=16)
abline(h = 0)


# residuals against Mix factor
boxplot(residuals(selection_mod)~ logit_mean2$mix, 
         varwidth = TRUE,
         ylab = "Residuals",
         xlab = "Mix",
         main = "")
abline(h = 0) # variance in residuals look similar between levels of "mix" !


hist(selection_mod$scaledResiduals,col='blue',xlab="residuals", main="Check Normality")

# residuals vs fitted
plot(residuals(selection_mod)~fitted(selection_mod)
      , col='blue'
      , pch=16)
abline(h = 0)


```
#### population level slope

```{r}
# population level slope deviation
# extract the modes of the random effects = population slope - overall slope
pop_re <- ranef(selection_mod)$`pop_id:mix` # extract the modes of the random effects
# extract random effect of mix
mix_re <- ranef(selection_mod)$mix %>%
  select(generation) %>%
  tibble::rownames_to_column("mix") %>%
  rename(mix_random_effect = generation)
# overall slope
s_overall <- fixef(selection_mod)["generation"]
# compute population slope
pop_slopes <- pop_re %>%
  tibble::rownames_to_column("pop_mix") %>%
  mutate(pop = substr(pop_mix, 1,3)) %>%
  mutate(mix = substr(pop_mix, 5,5))%>%
  rename(pop_random_effect = generation) %>%
  left_join(mix_re, by = "mix") %>%
  mutate(
    selection_coef = s_overall + pop_random_effect + mix_random_effect)%>%
  left_join(pop_info[, c(3,4)], by = c("pop" = "pop_id"))

```

# Plotting population-level selection coefficients

```{r, fig.width=8, fig.height=5}
mix_levels <- levels(as.factor(pop_slopes$mix))
strip_cols <- mix_colors[seq_along(mix_levels)]

ggplot( pop_slopes, aes(x = pop, y = selection_coef)) +
  geom_hline( yintercept = s_overall, linetype = "dashed", color = "grey40", 
              linewidth = 1) +
  geom_hline(yintercept=0, linetype = "solid") +
  geom_segment( aes(x = pop, xend = pop,
                    y = selection_coef, yend = s_overall,color = mix),
                linetype = "dashed", linewidth = 1, alpha = 0.7) +
  geom_point( aes(color = mix), size = 3, alpha = 0.7) +
  scale_color_manual(values = mix_colors)+
  theme_classic() +
  #facet_nested(~ mix, scales = "free_x", space = "free_x",
               #strip =strip_themed( background_x = elem_list_rect(
               #fill = scales::alpha(strip_cols, 0.7), 
               #colour = strip_cols)))+
  xlab("Population") +
  ylab("Selection coefficient") +
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 9),
    axis.text.x = element_text( angle = 45, hjust = 1, vjust = 1
    )
  )
```

## fig3 population slope distribution

```{r}
# set position of overall point
dens <- density(pop_slopes$selection_coef, na.rm = TRUE)
y_s <- max(dens$y) * 0.15

ggplot(pop_slopes, aes(x = selection_coef)) +
  geom_density( fill = "grey", alpha = 0.5, color = "black", linewidth = 1) +
  geom_vline( xintercept = s_overall, linetype = "dashed", linewidth = 1) +
  geom_errorbarh(aes( xmin = ci_s[1], xmax = ci_s[2],  y = y_s),
                 inherit.aes = FALSE, height = 2, linewidth = 1) +
  geom_point( aes( x = s_overall, y = y_s), 
              inherit.aes = FALSE, size = 3 ) +
  theme_classic() +
  theme(
  axis.title = element_text(size = 15),
  axis.text = element_text(size = 12))+
  xlab("Population-specific selection coefficient") +
  ylab("Density")
```


